\textbf{Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces}

### 2020-05-29 submission/manuscript.Rmd
Line break optimization.

### Illumina 16S rRNA sequencing
Illumina MiSeq sequencing of the V4 region of the 16S rRNA gene was performed as described previously [@Korlevica, submitted]. The V4 region of the 16S rRNA gene was amplified using a two-step PCR procedure. In the first PCR, the 515F (5'-GTGYCAGCMGCCGCGGTAA-3') and 806R (5'-GGACTACNVGGGTWTCTAAT-3') primers from the Earth Microbiome Project (http://press.igsb.anl.gov/earthmicrobiome/protocols-and-standards/16s/) were used [@Caporaso2012; @Apprill2015; @Parada2016]. These primers contained on their 5' end a tagged sequence. Purified PCR products were sent for Illumina MiSeq sequencing at IMGM Laboratories, Martinsried, Germany. Prior to sequencing at IMGM, the second PCR amplification of the two-step PCR procedure was performed using primers targeting the tagged region incorporated in the first PCR. In addition, these primers contained adapter and sample-specific index sequences. Beside samples, a positive and negative control for each sequencing batch was sequenced. The negative control comprised PCR reactions without DNA template, while for a positive control a mock community composed of evenly mixed DNA material originating from 20 bacterial strains (ATCC MSA-1002, ATCC, USA) was used. Sequences obtained in this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB37267.

>>>>

### Illumina 16S rRNA sequencing
Illumina MiSeq sequencing of the V4 region of the 16S rRNA gene was performed as described previously [@Korlevica, submitted]. The V4 region of the 16S rRNA gene was amplified using a two-step PCR procedure. In the first PCR, the 515F\linebreak(5'-GTGYCAGCMGCCGCGGTAA-3') and 806R (5'-GGACTACNVGGGTWTCTAAT-3') primers from the Earth Microbiome Project (http://press.igsb.anl.gov/earthmicrobiome/protocols-and-standards/16s/) were used [@Caporaso2012; @Apprill2015; @Parada2016]. These primers contained on their 5' end a tagged sequence. Purified PCR products were sent for Illumina MiSeq sequencing at IMGM Laboratories, Martinsried, Germany. Prior to sequencing at IMGM, the second PCR amplification of the two-step PCR procedure was performed using primers targeting the tagged region incorporated in the first PCR. In addition, these primers contained adapter and sample-specific index sequences. Beside samples, a positive and negative control for each sequencing batch was sequenced. The negative control comprised PCR reactions without DNA template, while for a positive control a mock community composed of evenly mixed DNA material originating from 20 bacterial strains (ATCC MSA-1002, ATCC, USA) was used. Sequences obtained in this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession number PRJEB37267.

### 2020-05-29 submission/manuscript.tex
Line break optimization.

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of sequences without
known relatives (no relative sequences) and eukaryotic, chloroplast and
mitochondrial sequences (\autoref{supp-nseq_notus}). A total of 35
samples originating from epiphytic archaeal and bacterial communities
associated with surfaces of the seagrass \emph{C. nodosa} and the
macroalga \emph{C. cylindracea} were analyzed. In addition, 18 samples
(one of the samples was sequenced twice) originating from the ambient
seawater were also processed for comparison. The number of reads per
sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off as commonly observed in
high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences as mentioned above reads were clustered into 28,750
different OTUs at a similarity level of 97 \si{\percent}. Read numbers
were normalized to the minimum number of sequences (8,408,
\autoref{supp-nseq_notus}) through rarefaction resulting in 17,036
different OTUs with 362 to 2,029 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes in richness
and diversity the observed number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,684.5 ±
131.3 OTUs{]} and monospecific {[}Number of OTUs, 1,727.6 ± 161.5
OTUs{]}) than for \emph{C. nodosa} (Number of OTUs, 1,060.2 ± 212.3
OTUs) and lowest values were obtained for the microbial community of the
ambient seawater (Number of OTUs, 524.5 ± 137.4 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not reveal such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly higher values
in spring and summer than in autumn and winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient, respectively, were calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from the
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared values were
calculated between ambient seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in-between the values of mixed
and monospecific \emph{C. cylindracea}. To assess seasonal changes in
the proportion of shared OTUs and communities the Jaccard's and
Bray-Curtis Similarity Coefficients were calculated between consecutive
sampling points (\autoref{shared}). Both coefficients showed similar
trends. Temporal proportional changes were more pronounced for ambient
seawater than for \emph{C. nodosa} and especially \emph{C. cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were
present at all seasons. These persistent OTUs constituted a high
proportion of total sequences (39.1 -- 52.4 \si{\percent}). To further
disentangle the environmental and seasonal community dissimilarity a
Principal Coordinates Analysis (PCoA) based on Bray-Curtis distances and
OTU abundances was applied. A clear separation between ambient seawater
and surface associated communities was found (\autoref{pcoa}). In
addition, a separation of epiphytic bacterial and archaeal communities
based on host species was detected. This separation was further
supported by ANOSIM (R = 0.96, \emph{p} \textless{} 0.001). Seasonal
changes of \emph{C. nodosa} associated communities indicated a
separation between spring, summer and autumn/winter samples (ANOSIM, R =
0.53, \emph{p} \textless{} 0.01). For \emph{C. cylindracea} associated
communities a separation between summer and autumn/winter/spring samples
was observed that was, however, not as strong as for \emph{C. nodosa}
associated communities (ANOSIM, R = 0.31, \emph{p} \textless{} 0.01)
(\autoref{pcoa}).

@@

Seasonal changes in richness in the epiphytic community were substantial
as indicated by the proportion of OTUs (\(\leq\) 1.0 \si{\percent})
present at every sampling date. These persistent OTUs, however, were
accounting for a high proportion of sequences (\(\leq\) 52.4
\si{\percent}) (\autoref{shared}). A very similar proportion of
persistent OTUs was reported in high-frequency sampling studies
describing seasonal changes in picoplankton (Gilbert \emph{et al.},
2009, 2012). In comparison to the seawater community, a lower degree of
seasonal shifts was observed for the macrophyte surface associated
communities. It appears that microniches at the surfaces of macrophytes
are providing more stable conditions than the ambient seawater. At the
level of OTUs seasonal changes of \emph{C. nodosa} and \emph{C.
cylindracea} associated communities were identified that could be linked
to the growth cycle of the seagrass and macroalgae (M. Najdek, personal
communication). \emph{C. nodosa} was characterized by a spring community
during maximum seagrass proliferation, a summer community during the
highest standing stock of \emph{C.nodosa} and an autumn/winter community
during the decay of seagrass biomass. In contrast, \emph{C. cylindracea}
started to proliferate in late spring and was characterized only by a
summer community during high growth rates and by an autumn/winter/spring
community when the biomass was at the peak and decaying thereafter.
Similar seasonal changes in the epiphytic community were also described
for other macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et al.},
2011). Higher seasonal stability of \emph{C. cylindracea} surface
communities than in \emph{C. nodosa} was also observed in the higher
proportion of shared communities between two consecutive sampling dates
in \emph{C. cylindracea}.

@@

Illumina MiSeq sequencing of the V4 region of the 16S rRNA gene was
performed as described previously (Korlević \emph{et al.}, submitted).
The V4 region of the 16S rRNA gene was amplified using a two-step PCR
procedure. In the first PCR, the 515F (5'-GTGYCAGCMGCCGCGGTAA-3') and
806R (5'-GGACTACNVGGGTWTCTAAT-3') primers from the Earth Microbiome
Project
(\url{http://press.igsb.anl.gov/earthmicrobiome/protocols-and-standards/16s/})
were used (Caporaso \emph{et al.}, 2012; Apprill \emph{et al.}, 2015;
Parada \emph{et al.}, 2016). These primers contained on their 5' end a
tagged sequence. Purified PCR products were sent for Illumina MiSeq
sequencing at IMGM Laboratories, Martinsried, Germany. Prior to
sequencing at IMGM, the second PCR amplification of the two-step PCR
procedure was performed using primers targeting the tagged region
incorporated in the first PCR. In addition, these primers contained
adapter and sample-specific index sequences. Beside samples, a positive
and negative control for each sequencing batch was sequenced. The
negative control comprised PCR reactions without DNA template, while for
a positive control a mock community composed of evenly mixed DNA
material originating from 20 bacterial strains (ATCC MSA-1002, ATCC,
USA) was used. Sequences obtained in this study have been deposited in
the European Nucleotide Archive (ENA) at EMBL-EBI under accession number
PRJEB37267.

>>>>

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of sequences without
known relatives (no relative sequences) and eukaryotic, chloroplast and
mitochondrial sequences (\autoref{supp-nseq_notus}). A total of 35
samples originating from epiphytic archaeal and bacterial communities
associated with surfaces of the seagrass \emph{C. nodosa} and the
macroalga \emph{C. cylindracea} were analyzed. In addition, 18 samples
(one of the samples was sequenced twice) originating from the ambient
seawater were also processed for comparison. The number of reads per
sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off as commonly observed in
high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences as mentioned above reads were clustered into 28,750
different OTUs at a similarity level of 97 \si{\percent}. Read numbers
were normalized to the minimum number of sequences (8,408,
\autoref{supp-nseq_notus}) through rarefaction resulting in 17,041
different OTUs with 359 to 1,995 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes in richness
and diversity the observed number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,698.7 ±
142.1 OTUs{]} and monospecific {[}Number of OTUs, 1,731.1 ± 151.9
OTUs{]}) than for \emph{C. nodosa} (Number of OTUs, 1,053.7 ± 209.6
OTUs) and lowest values were obtained for the microbial community of the
ambient seawater (Number of OTUs, 530.6 ± 144.6 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not reveal such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly higher values
in spring and summer than in autumn and winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient, respectively, were calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from the
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.78), while lower shared values were
calculated between ambient seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in-between the values of mixed
and monospecific \emph{C. cylindracea}. To assess seasonal changes in
the proportion of shared OTUs and communities the Jaccard's and
Bray-Curtis Similarity Coefficients were calculated between consecutive
sampling points (\autoref{shared}). Both coefficients showed similar
trends. Temporal proportional changes were more pronounced for ambient
seawater than for \emph{C. nodosa} and especially \emph{C. cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were
present at all seasons. These persistent OTUs constituted a high
proportion of total sequences (40.6 -- 53.1 \si{\percent}). To further
disentangle the environmental and seasonal community dissimilarity a
Principal Coordinates Analysis (PCoA) based on Bray-Curtis distances and
OTU abundances was applied. A clear separation between ambient seawater
and surface associated communities was found (\autoref{pcoa}). In
addition, a separation of epiphytic bacterial and archaeal communities
based on host species was detected. This separation was further
supported by ANOSIM (R = 0.96, \emph{p} \textless{} 0.001). Seasonal
changes of \emph{C. nodosa} associated communities indicated a
separation between spring, summer and autumn/winter samples (ANOSIM, R =
0.53, \emph{p} \textless{} 0.01). For \emph{C. cylindracea} associated
communities a separation between summer and autumn/winter/spring samples
was observed that was, however, not as strong as for \emph{C. nodosa}
associated communities (ANOSIM, R = 0.33, \emph{p} \textless{} 0.01)
(\autoref{pcoa}).

@@

Seasonal changes in richness in the epiphytic community were substantial
as indicated by the proportion of OTUs (\(\leq\) 1.0 \si{\percent})
present at every sampling date. These persistent OTUs, however, were
accounting for a high proportion of sequences (\(\leq\) 53.1
\si{\percent}) (\autoref{shared}). A very similar proportion of
persistent OTUs was reported in high-frequency sampling studies
describing seasonal changes in picoplankton (Gilbert \emph{et al.},
2009, 2012). In comparison to the seawater community, a lower degree of
seasonal shifts was observed for the macrophyte surface associated
communities. It appears that microniches at the surfaces of macrophytes
are providing more stable conditions than the ambient seawater. At the
level of OTUs seasonal changes of \emph{C. nodosa} and \emph{C.
cylindracea} associated communities were identified that could be linked
to the growth cycle of the seagrass and macroalgae (M. Najdek, personal
communication). \emph{C. nodosa} was characterized by a spring community
during maximum seagrass proliferation, a summer community during the
highest standing stock of \emph{C.nodosa} and an autumn/winter community
during the decay of seagrass biomass. In contrast, \emph{C. cylindracea}
started to proliferate in late spring and was characterized only by a
summer community during high growth rates and by an autumn/winter/spring
community when the biomass was at the peak and decaying thereafter.
Similar seasonal changes in the epiphytic community were also described
for other macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et al.},
2011). Higher seasonal stability of \emph{C. cylindracea} surface
communities than in \emph{C. nodosa} was also observed in the higher
proportion of shared communities between two consecutive sampling dates
in \emph{C. cylindracea}.

@@

Illumina MiSeq sequencing of the V4 region of the 16S rRNA gene was
performed as described previously (Korlević \emph{et al.}, submitted).
The V4 region of the 16S rRNA gene was amplified using a two-step PCR
procedure. In the first PCR, the
515F\linebreak(5'-GTGYCAGCMGCCGCGGTAA-3') and 806R
(5'-GGACTACNVGGGTWTCTAAT-3') primers from the Earth Microbiome Project
(\url{http://press.igsb.anl.gov/earthmicrobiome/protocols-and-standards/16s/})
were used (Caporaso \emph{et al.}, 2012; Apprill \emph{et al.}, 2015;
Parada \emph{et al.}, 2016). These primers contained on their 5' end a
tagged sequence. Purified PCR products were sent for Illumina MiSeq
sequencing at IMGM Laboratories, Martinsried, Germany. Prior to
sequencing at IMGM, the second PCR amplification of the two-step PCR
procedure was performed using primers targeting the tagged region
incorporated in the first PCR. In addition, these primers contained
adapter and sample-specific index sequences. Beside samples, a positive
and negative control for each sequencing batch was sequenced. The
negative control comprised PCR reactions without DNA template, while for
a positive control a mock community composed of evenly mixed DNA
material originating from 20 bacterial strains (ATCC MSA-1002, ATCC,
USA) was used. Sequences obtained in this study have been deposited in
the European Nucleotide Archive (ENA) at EMBL-EBI under accession number
PRJEB37267.
